Lactate transport inhibition therapeutically reprograms fibroblast

### 2024-04-25 manuscript/manuscript.qmd
updates in response to co-author feedback

authors:
- name: David R. Ziehr
  affiliations: 
  - ref: bwh
  - ref: mgh
  - ref: hms
- name: Fei Li
  affiliations:
  - ref: bwh
  - ref: hms
- name: K. Mark Parnell
  affiliations:
  - ref: vettore
- name: Nathan M. Krah
  affiliations:
  - ref: utah_gen
- name: Kevin J. Leahy 
  affiliations:
  - ref: bwh
- name: Christelle Guillermier
  affiliations:
  - ref: bwh
  - ref: hms
- name: Jack Varon
  affiliations: 
  - ref: bwh
  - ref: hms
- name: Rebecca M. Baron
  affiliations:
  - ref: bwh
  - ref: hms
- name: Bradley A. Maron
  affiliations:
  - ref: umd
  - ref: ihc
- name: Nancy J. Philp
  affiliations:
  - ref: tj
- name: Lida P. Hariri
  affiliations: 
  - ref: hms
  - ref: mgh_path
- name: Edy Y. Kim
  affiliations:
  - ref: bwh
  - ref: hms
- name: Matthew L. Steinhauser
  affiliations:
  - ref: pitt
  - ref: upmc
- name: Rachel S. Knipe
  affiliations:
  - ref: mgh
  - ref: hms
- name: Jared Rutter
  affiliations:
  - ref: utah_biochem
  - ref: utah_hhmi
- name: William M. Oldham
  affiliations:
  - ref: bwh
  - ref: hms
  email: woldham@bwh.harvard.edu
  corresponding: true

affiliations:
- id: bwh
  name: Department of Medicine, Brigham and Women's Hospital, Boston, MA
- id: mgh
  name: Department of Medicine, Massachusetts General Hospital, Boston, MA
- id: hms
  name: Department of Medicine, Harvard Medical School, Boston, MA
- id: vettore
  name: Vettore Biosciences, San Francisco, CA
- id: utah_gen
  name: Department of Internal Medicine, University of Utah, Salt Lake City, UT
- id: umd
  name: Department of Medicine, University of Maryland School of Medicine, Baltimore, MA
- id: ihc
  name: University of Maryland Institute for Health Computing, Bethesda, MD
- id: tj
  name: Department of Pathology and Genomic Medicine, Thomas Jefferson University, Philadelphia, PA
- id: mgh_path
  name: Department of Pathology, Massachusetts General Hospital, Boston, MA
- id: pitt
  name: Aging Institute, University of Pittsburgh, Pittsburgh, PA
- id: upmc
  name: UPMC Heart and Vascular Institute, UPMC Presbyterian, Pittsburgh, PA
- id: utah_biochem
  name: Department of Biochemistry, University of Utah, Salt Lake City, UT
- id: utah_hhmi
  name: Howard Hughes Medical Institute, University of Utah School of Medicine, Salt Lake City, UT

@@

Small molecule inhibitors of lactate transporters, including the novel MCT4 inhibitor VB253, reprogram fibroblast metabolism to prevent myofibroblast differentiation and to decrease bleomycin-induced pulmonary fibrosis.

@@

Myofibroblast differentiation, essential for driving extracellular matrix synthesis in pulmonary fibrosis, requires increased glycolysis. While glycolytic cells must export lactate, the contributions of lactate transporters to myofibroblast differentiation are unknown. In this study, we investigated how MCT1 and MCT4, key lactate transporters, influence myofibroblast differentiation and experimental pulmonary fibrosis. Our findings reveal that inhibiting MCT1 or MCT4 reduces TGFβ-stimulated myofibroblast differentiation *in vitro* and decreases bleomycin-induced pulmonary fibrosis *in vivo*. Through comprehensive metabolic analyses, including bioenergetics, stable isotope tracing, metabolomics, and imaging mass spectrometry in both cells and mice, we demonstrate that inhibiting lactate transport enhances oxidative phosphorylation, reduces reactive oxygen species production, and diminishes glucose metabolite incorporation into fibrotic lung regions. Furthermore, we introduce VB253, a novel MCT4 inhibitor, which effectively ameliorated pulmonary fibrosis in both young and aged mice, comparable to established antifibrotic therapies. These results underscore the necessity of lactate transport for myofibroblast differentiation, identify MCT1 and MCT4 as promising pharmacologic targets in pulmonary fibrosis, and support further evaluation of lactate transport inhibitors for patients with pulmonary fibrosis for whom limited therapeutic options currently exist.

@@

Accumulating evidence suggests that metabolic reprogramming may be one such therapeutic vulnerability in IPF [@bueno2020; @selvarajah2021]. Lung fibrosis is driven by the excessive deposition of extracellular matrix by myofibroblasts [@martinez2017]. Fundamental changes in myofibroblast metabolism support myofibroblast differentiation and extracellular matrix production [@bernard2015; @xie2015; @goodwin2018; @hamanaka2019; @selvarajah2019]. In particular, increased glycolysis and lactate production have been observed in IPF myofibroblasts *ex vivo* and following transforming growth factor β1 (TGFβ)-induced myofibroblast differentiation *in vitro* [@kottmann2012; @xie2015; @kottmann2015; @bernard2015]. These metabolic changes are critical for fibrogenesis, as small molecule inhibitors of glucose uptake, glycolysis, and lactate fermentation prevent myofibroblast differentiation *in vitro* and attenuate pulmonary fibrosis in animal models [@bernard2015; @kottmann2012; @xie2015; @goodwin2018; @judge2018; @cho2017]. Unfortunately, low target affinities, poor specificity, narrow therapeutic indices, and common genetic resistance have all hampered the translation of these investigational compounds for clinical use [@rodriguez-enriquez2009; @michelakis2017; @pelicano2006]. Moreover, the molecular mechanisms by which these metabolic inhibitors attenuate the myofibroblast differentiation transcriptional program remain unclear. In order to leverage metabolic therapies for IPF, more targeted and better characterized drugs must be developed.

Toward this end, we aimed to examine the impact of a novel metabolic strategy --- lactate transport inhibition --- on myofibroblast differentiation and experimental pulmonary fibrosis. Myofibroblasts rely on lactate secretion to sustain glycolysis, a process regulated by a family of monocarboxylate transporters (MCT1-4). Inhibitors targeting these transporters have been actively explored in clinical trials for oncological conditions where glycolytic reprogramming also features prominently in disease pathobiology [@puri2020; @cluntun2021]. Importantly, MCT inhibitors present favorable pharmacologic profiles compared to previously studied glycolysis inhibitors, with successful translation to human clinical trials for advanced solid tumors [@halford2023]. Before the promise of this therapeutic approach in IPF may be realized, however, the preclinical efficacy and molecular mechanisms-of-action of lactate transport inhibitors must be demonstrated experimentally.

In this work, we evaluated the contribution of lactate transporters to experimental pulmonary fibrosis. We found increased expression of the lactate transporters MCT1 and MCT4 IPF patient lungs. Inhibition of these transporters attenuated bleomycin-induced lung fibrosis *in vivo* and TGFβ-induced myofibroblast differentiation *in vitro*, where MCT4 inhibition demonstrated increased therapeutic efficacy. Using metabolomics, stable isotope tracing, and high-resolution spatial metabolomic imaging, we find that lactate transport inhibition promotes glucose oxidation and decreases pro-fibrogenic reactive oxygen species (ROS) production. Building on these results, we introduce VB253, a novel MCT4 inhibitor suitable for human clinical studies. Together, our findings offer insights into disrupting the metabolic pathways driving IPF fibrosis and present a promising strategy targeting lactate transporters for the treatment of this fatal condition.

@@

We proceeded to investigate whether MCT expression and activity were essential for myofibroblast differentiation *in vitro* using RNA interference and pharmacologic approaches. Lung fibroblasts were transfected with siRNA targeting MCT1 and MCT4 individually and in combination. After 24 h, the cells were treated with TGFβ for 48 h to induce myofibroblast differentiation. The siRNAs effectively reduced lactate transporter protein levels (**Fig. 2A**). Reduction in either MCT1 or MCT4 expression caused a marked decrease in TGFβ-stimulated α-SMA expression. Notably, siMCT1 also decreased MCT4 expression. No adverse effects on cell viability were noted following lactate transporter knockdown, and siMCT1 significantly increased cell count in both control and TGFβ-treated cells (**Fig. S1A**).

@@

MCTs play pivotal roles in maintaining cellular lactate homeostasis. MCT1 serves as the primary lactate importer in cells utilizing lactate for oxidative phosphorylation or gluconeogenesis and exhibits ubiquitous expression. Additionally, MCT1 facilitates lactate export in glycolytic cells [@halestrap2013; @brooks2018]. By contrast, MCT4 functions as the main lactate exporter in glycolytic cells and is up-regulated by hypoxia-inducible transcription factors [@ullah2006; @halestrap2013]. Importantly, MCT4 can also act as a lactate importer with a *K*~M~ of 1 mM [@contreras-baeza2019].

To assess the metabolic consequences of lactate transporter inhibition, we quantified extracellular lactate in the conditioned medium from cells treated with MCT siRNA or pharmacologic inhibitors. Silencing MCT1 and MCT4, either individually or concurrently, decreased TGFβ-stimulated lactate efflux in lung fibroblasts (**Fig. 3A**). By contrast, pharmacologic inhibition of either MCT1 or MCT4 alone did not decrease TGFβ-stimulated lactate efflux (**Fig. 3B**); inhibition of both transporters was required to lower extracellular lactate levels. Inhibition of MCT1, MCT2, and MCT4 with the combination of AR-C155858 and VB124 was required to prevent increases in lactate efflux (**Fig. S3A**). Although less pronounced compared to lactate production, we observed a trend toward decreased extracellular glucose consumption in cells treated with both AZD3965 and VB124 (**Fig. S3B**). These findings align with previous studies indicating compensatory roles for MCT1 and MCT4 in lactate export [@benjamin2018]. Furthermore, these results suggest that glycolysis inhibition is not the principal mechanism underlying the attenuation of myofibroblast differentiation by lactate transport inhibition.

To further elucidate the metabolic consequences of lactate transport inhibition, we next measured proton efflux (PER) and oxygen consumption (OCR) rates in lung fibroblasts treated with TGFβ in combination with MCT inhibitors (**Fig. 3C, S3C**). Given that MCTs co-transport protons with lactate, PER serves as a surrogate measure of lactate efflux. In line with previous findings [@schruf2019; @cho2017; @xie2015; @bernard2015; @goodwin2018], we observed increases in both PER and OCR following 48 h of TGFβ stimulation, indicative of increased glycolysis and oxidative phosphorylation. Consistent with our direct measures of extracellular lactate and glucose, concurrent treatment with AZD3965 and VB124 was necessary to reduce PER. Consequently, this reduction in PER coincided with a corresponding increase in OCR as myofibroblasts transitioned their metabolism from glycolysis to oxidative phosphorylation.

@@

To provide further support for this hypothesis, we performed liquid chromatography-mass spectrometry-based profiling of extracellular and intracellular metabolites from cells treated with AZD3965 and VB124. In accordance with the Seahorse analysis, inhibiting a single lactate transporter had modest effects on extracellular metabolite levels (**Fig. 4A-D, S4**). This analysis confirmed the results of extracellular lactate measurements by enzyme assay, which showed that dual inhibition was required to decrease lactate efflux (**Fig. 4B**). Besides lactate, dual inhibition of lactate transporters primarily altered transcellular fluxes of amino acids (**Fig. 4C-D**), including several metabolites that were differentially regulated by TGFβ treatment (**Fig. S4A**) and dual lactate transport inhibition, including leucine, alanine, ornithine, and ketoleucine.

Similar effects were observed on intracellular metabolites. PCA showed distinct clustering of treatment groups, with the drug effects aligning with PC1 and TGFβ treatment effects with PC2 (**Fig. 4E**). Similar to the extracellular flux results, the magnitude of drug-induced perturbations was increased from MCT1 to MCT4 to combined inhibition (**Fig. 4E**), reflected by intracellular lactate levels (**Fig. 4F**). As expected from extracellular lactate measures following MCT inhibition, intracellular lactate accumulation moderately increased with MCT4 inhibition and substantially with MCT1/4 inhibition. TGFβ treatment resulted in diverse changes in the intracellular metabolomic profile of treated fibroblasts (**Fig. S4C-D**). The effect of AZD3965 alone on intracellular metabolite levels was modest (**Fig. S4G-H**). Conversely, MCT4 inhibition by VB124 alone (**Fig. SK-L**) or in combination with AZD3965 (**Fig. 4G-H**) caused substantial perturbations to intracellular metabolism. Specifically, we observed enrichment of the glycolysis and tricarboxylic acid (TCA) cycle metabolite sets with MCT4 inhibition. These findings collectively suggest that inhibiting lactate export leads to the accumulation of upstream glycolytic intermediates that are rerouted to mitochondrial oxidative metabolic pathways.

To test this hypothesis, we labeled lung fibroblasts with [U-^13^C~6~]-glucose (8 mM) in medium containing lactate (2 mM), glutamine (1 mM), and pyruvate (1 mM) during TGFβ stimulation and treatment with MCT inhibitors (**Fig. 4J, S5**). TGFβ increases ^13^C incorporation from glucose into pyruvate, lactate, citrate, succinate (SUC), and malate (MAL), indicated enhanced flux from glucose into the TCA cycle. While AZD3965 had minimal impact on these labeling patterns, MCT4 inhibition significantly elevated the fractions of these metabolites labeled by glucose, providing direct evidence for a proportional increase in glucose oxidation following MCT4 inhibition, consistent with the bioenergetic and steady-state metabolomics experiments described above.

Increased metabolite labeling by glucose must by offset by decreased labeling from other substrates. Given recent data suggesting lactate as a major oxidative fuel source in the lung [@hui2017; @hui2020; @copeland2023], we hypothesized that MCT inhibition would decrease exogenous lactate oxidation. To test this, lung fibroblasts were cultured with [U-^13^C~3~]-lactate (2 mM) in medium containing naturally labeled glucose, glutamine, and pyruvate (**Fig. 4K, S5**). Extracellular [U-^13^C~3~]-lactate labeled approximately 50% of intracellular pyruvate and lactate at baseline, with significant downstream incorporation into TCA metabolites. This labeling decreased following TGFβ treatment, mirroring increased fractional labeling from glucose (**Fig. 4J**). MCT inhibition had no impact on fractional labeling of TCA intermediates by [U-^13^C]-glutamine (**Fig. S4M, S5**). Interestingly, MCT4 inhibition alone or in combination with AZD3965 decreased ^13^C labeling of intracellular metabolites by lactate, underscoring the importance of bi-directional lactate transport mediated by MCTs. Consistent with our measurements of extracellular lactate, the effects of MCT inhibition were more pronounced when both inhibitors were used simultaneously, again highlighting functional redundancy of MCT1 and MCT4 in lung fibroblasts.

@@

Given that increased intracellular lactate correlated with more potent inhibition of α-SMA expression, we treated cells with TGFβ in combination with extracellular lactate (10 mM). A previous study suggested that extracellular lactate modestly increased α-SMA expression independently of TGFβ [@kottmann2012]. However, we observed no impact of extracellular lactate on α-SMA expression in TGFβ-treated lung fibroblasts (**Fig. S7A**), indicating intracellular lactate accumulation alone does not mediate the antifibrotic effects of MCT inhibition.

@@

Our preclinical findings indicate that MCT4 inhibition, as a single therapeutic target, exhibits greater antifibrotic efficacy than MCT1 inhibition, both in suppressing myofibroblast differentiation *in vitro* and reducing bleomycin-induced fibrosis. VB253 is a novel inhibitor of MCT4 with similar selectivity, but ~10-fold increased potency, for MCT4 inhibition. Like VB124, VB253 dose-dependently decreases TGFβ-stimulated α-SMA in human IPF lung fibroblasts *ex vivo* (**Fig. 8A**). To assess its efficacy compared to established therapies, we compared VB253 to nintedanib, a clinically approved antifibrotic medication for pulmonary fibrosis (**Fig. 8A-B**). Nintedanib inhibits TGFβ-mediated myofibroblast differentiation and decreases Col1a1 expression *in vitro* and *in vivo*  [@wollin2014]. While both compounds effectively attenuated α-SMA expression, nintedanib exhibited moderate cytotoxicity absent with VB253 (**Fig. 8B**). Nintedanib mitigates myofibroblast differentiation partly by inhibiting TGFβ receptor phosphorylation and Smad-dependent signaling pathways [@rangarajan2016]. Correspondingly, nintedanib dose-dependently attenuated TGFβ-mediated Smad3 phosphorylation, while VB253 and VB124 have no impact on this signaling pathway (**Fig. 5, S10**).

Subsequently, we evaluated the potential of VB253 to counter bleomycin-induced pulmonary fibrosis. In experiments conducted independently from those reported above, VB253 was administered to young mice (8-10 weeks) beginning on day 7 following bleomycin (**Fig. 8C**). Nintedanib and pirfenidone served as comparators to existing pulmonary fibrosis therapies. Three weeks after bleomycin, whole-body plethysmography was performed to assess breathing patterns in unrestrained mice. Enhanced pause (Penh) is a dimensionless index describing airflow during tidal breathing found to increase significantly in bleomycin-treated mice [@vanoirbeek2010]. VB253 restored Penh to baseline levels, suggesting normalization of respiratory patterns in bleomycin-treated mice (**Fig. 8D**). Histologic assessment confirmed reduced pulmonary fibrosis (**Fig. 8E**) and α-SMA expression (**Fig. 8F**). Moreover, the effects of VB253 were comparable to the antifibrotic effects of nintedanib and pirfenidone with decreased cytotoxicity in *in vitro* assays.

@@

This study was designed to investigate the role of lactate transporters in the myofibroblast differentiation and pulmonary fibrosis. The objectives of this study were (i) to determine the expression of lactate transporters in IPF lungs and model systems, (ii) to characterize the phenotypic effects of lactate transport inhibition in model systems, and (iii) to profile the metabolic consequences of lactate transport inhibitors.

>>>>

authors:
- name: David R. Ziehr
  affiliations: 
  - ref: bwh
  - ref: mgh
  - ref: hms
- name: Fei Li
  affiliations:
  - ref: bwh
  - ref: hms
- name: K. Mark Parnell
  affiliations:
  - ref: vettore
- name: Nathan M. Krah
  affiliations:
  - ref: utah_gen
  - ref: utah_biochem
- name: Kevin J. Leahy 
  affiliations:
  - ref: bwh
- name: Christelle Guillermier
  affiliations:
  - ref: bwh
  - ref: hms
- name: Jack Varon
  affiliations: 
  - ref: bwh
  - ref: hms
- name: Rebecca M. Baron
  affiliations:
  - ref: bwh
  - ref: hms
- name: Bradley A. Maron
  affiliations:
  - ref: umd
  - ref: ihc
- name: Nancy J. Philp
  affiliations:
  - ref: tj
- name: Lida P. Hariri
  affiliations: 
  - ref: hms
  - ref: mgh_path
- name: Edy Y. Kim
  affiliations:
  - ref: bwh
  - ref: hms
- name: Matthew L. Steinhauser
  affiliations:
  - ref: pitt
  - ref: upmc
- name: Rachel S. Knipe
  affiliations:
  - ref: mgh
  - ref: hms
- name: Jared Rutter
  affiliations:
  - ref: utah_biochem
  - ref: utah_hhmi
- name: William M. Oldham
  affiliations:
  - ref: bwh
  - ref: hms
  email: woldham@bwh.harvard.edu
  corresponding: true

affiliations:
- id: bwh
  name: Department of Medicine, Brigham and Women's Hospital, Boston, MA
- id: mgh
  name: Department of Medicine, Massachusetts General Hospital, Boston, MA
- id: hms
  name: Department of Medicine, Harvard Medical School, Boston, MA
- id: vettore
  name: Vettore Biosciences, San Francisco, CA
- id: utah_gen
  name: Department of Internal Medicine, University of Utah, Salt Lake City, UT
- id: utah_biochem
  name: Department of Biochemistry, University of Utah, Salt Lake City, UT
- id: umd
  name: Department of Medicine, University of Maryland School of Medicine, Baltimore, MD
- id: ihc
  name: University of Maryland Institute for Health Computing, Bethesda, MD
- id: tj
  name: Department of Pathology and Genomic Medicine, Thomas Jefferson University, Philadelphia, PA
- id: mgh_path
  name: Department of Pathology, Massachusetts General Hospital, Boston, MA
- id: pitt
  name: Aging Institute, University of Pittsburgh, Pittsburgh, PA
- id: upmc
  name: UPMC Heart and Vascular Institute, UPMC Presbyterian, Pittsburgh, PA
- id: utah_hhmi
  name: Howard Hughes Medical Institute, University of Utah School of Medicine, Salt Lake City, UT

@@

Small molecule inhibitors of lactate transporters, including the novel MCT4 inhibitor VB253, reprogram fibroblast metabolism to prevent myofibroblast differentiation and decrease bleomycin-induced pulmonary fibrosis.

@@

Myofibroblast differentiation, essential for driving extracellular matrix synthesis in pulmonary fibrosis, requires increased glycolysis. While glycolytic cells must export lactate, the contributions of lactate transporters to myofibroblast differentiation are unknown. In this study, we investigated how MCT1 and MCT4, key lactate transporters, influence myofibroblast differentiation and experimental pulmonary fibrosis. Our findings reveal that inhibiting MCT1 or MCT4 reduces TGFβ-stimulated pulmonary myofibroblast differentiation *in vitro* and decreases bleomycin-induced pulmonary fibrosis *in vivo*. Through comprehensive metabolic analyses, including bioenergetics, stable isotope tracing, metabolomics, and imaging mass spectrometry in both cells and mice, we demonstrate that inhibiting lactate transport enhances oxidative phosphorylation, reduces reactive oxygen species production, and diminishes glucose metabolite incorporation into fibrotic lung regions. Furthermore, we introduce VB253, a novel MCT4 inhibitor, which ameliorates pulmonary fibrosis in both young and aged mice, with comparable efficacy to established antifibrotic therapies. These results underscore the necessity of lactate transport for myofibroblast differentiation, identify MCT1 and MCT4 as promising pharmacologic targets in pulmonary fibrosis, and support further evaluation of lactate transport inhibitors for patients for whom limited therapeutic options currently exist.

@@

Accumulating evidence suggests that metabolic reprogramming may be one such therapeutic strategy in IPF [@bueno2020; @selvarajah2021]. Lung fibrosis is driven by the excessive deposition of extracellular matrix by myofibroblasts [@martinez2017]. Fundamental changes in myofibroblast metabolism support myofibroblast differentiation and extracellular matrix production [@bernard2015; @xie2015; @goodwin2018; @hamanaka2019; @selvarajah2019]. In particular, increased glycolysis and lactate production have been observed in IPF myofibroblasts *ex vivo* and following transforming growth factor β1 (TGFβ)-induced myofibroblast differentiation *in vitro* [@kottmann2012; @xie2015; @kottmann2015; @bernard2015]. These metabolic changes are critical for fibrogenesis, as small molecule inhibitors of glucose uptake, glycolysis, and lactate fermentation prevent myofibroblast differentiation *in vitro* and attenuate pulmonary fibrosis in animal models [@bernard2015; @kottmann2012; @xie2015; @goodwin2018; @judge2018; @cho2017]. Unfortunately, low target affinities, poor specificity, narrow therapeutic indices, and common genetic resistance have all hampered the translation of these investigational compounds for clinical use [@rodriguez-enriquez2009; @michelakis2017; @pelicano2006]. Moreover, the molecular mechanisms by which these metabolic inhibitors attenuate the myofibroblast differentiation transcriptional program remain unclear. In order to leverage metabolic therapies for IPF, more targeted and better characterized drugs must be developed.

Toward this end, we aimed to examine the impact of a novel metabolic strategy --- lactate transport inhibition --- on myofibroblast differentiation and experimental pulmonary fibrosis. Sustained glycolysis in myofibroblasts relies on lactate secretion, which is conducted by a family of monocarboxylate transporters (MCT1-4). Inhibitors targeting these transporters have been actively explored in clinical trials for oncological conditions where glycolytic reprogramming also features prominently in disease pathobiology [@puri2020; @cluntun2021]. Importantly, MCT inhibitors present favorable pharmacologic profiles compared to previously studied glycolysis inhibitors, with successful translation to human clinical trials for advanced solid tumors [@halford2023]. Before the promise of this therapeutic approach in IPF may be realized, however, the preclinical efficacy and molecular mechanisms-of-action of lactate transport inhibitors must be demonstrated experimentally.

In this work, we evaluated the contribution of lactate transporters to experimental pulmonary fibrosis. We found increased expression of the lactate transporters MCT1 and MCT4 IPF patient lungs. Inhibition of these transporters attenuated bleomycin-induced lung fibrosis *in vivo* and TGFβ-induced myofibroblast differentiation *in vitro*, where MCT4 inhibition demonstrated increased therapeutic efficacy. Using metabolomics, stable isotope tracing, and high-resolution spatial metabolomic imaging, we find that lactate transport inhibition promotes glucose oxidation and decreases pro-fibrogenic reactive oxygen species (ROS) production. Building on these results, we introduce VB253, a novel MCT4 inhibitor suitable for human clinical studies that performs similarly to current standard-of-care antifibrotic therapies in bleomycin-induced pulmonary fibrosis. Together, our findings offer insights into disrupting the metabolic pathways driving IPF fibrosis and present a promising strategy targeting lactate transporters for the treatment of this fatal condition.

@@

We proceeded to investigate whether MCT expression and activity were essential for myofibroblast differentiation *in vitro* using RNA interference and pharmacologic approaches. Lung fibroblasts were transfected with siRNA targeting MCT1 and MCT4 individually and in combination. After 24 h, the cells were treated with TGFβ for 48 h to induce myofibroblast differentiation. The siRNAs reduced lactate transporter protein levels (**Fig. 2A**). Reduction in either MCT1 or MCT4 expression caused a marked decrease in TGFβ-stimulated α-SMA expression. Notably, siMCT1 also decreased MCT4 expression. No adverse effects on cell viability were noted following lactate transporter knockdown, and siMCT1 significantly increased cell count in both control and TGFβ-treated cells (**Fig. S1A**).

@@

MCTs play pivotal roles in maintaining cellular lactate and redox homeostasis. MCT1 predominantly imports lactate in cells utilizing lactate for oxidative phosphorylation or gluconeogenesis and is ubiquitously expressed. MCT1 also facilitates lactate export in some glycolytic cells [@halestrap2013; @brooks2018]. By contrast, MCT4 functions as the main lactate exporter in glycolytic cells and is up-regulated when the glycolytic transcriptional program is activated by, for example, the c-Myc or hypoxia-inducible transcription factors [@ullah2006; @halestrap2013]. Importantly, MCT4 can also act as a lactate importer with a *K*~M~ of 1 mM [@contreras-baeza2019].

To assess the metabolic consequences of lactate transporter inhibition, we quantified extracellular lactate in the conditioned medium from cells treated with MCT siRNA or pharmacologic inhibitors. Silencing MCT1 and MCT4, either individually or concurrently, decreased net TGFβ-stimulated lactate efflux in lung fibroblasts (**Fig. 3A**). By contrast, pharmacologic inhibition of either MCT1 or MCT4 alone did not decrease TGFβ-stimulated lactate efflux (**Fig. 3B**); inhibition of both transporters was required to lower extracellular lactate levels. Inhibition of MCT1, MCT2, and MCT4 with the combination of AR-C155858 and VB124 was required to prevent increases in lactate efflux (**Fig. S3A**). Although less pronounced compared to lactate production, we observed a trend toward decreased extracellular glucose consumption in cells treated with both AZD3965 and VB124 (**Fig. S3B**). These findings align with previous studies indicating compensatory roles for MCT1 and MCT4 in lactate export [@benjamin2018]. Furthermore, these results suggest that glycolysis inhibition is not the principal mechanism underlying the attenuation of myofibroblast differentiation by lactate transport inhibition.

To further elucidate the metabolic consequences of lactate transport inhibition, we next measured proton efflux (PER) and oxygen consumption (OCR) rates in lung fibroblasts treated with TGFβ in combination with MCT inhibitors (**Fig. 3C, S3C**). Given that MCTs co-transport protons with lactate, PER serves as a surrogate measure of lactate efflux. In line with previous findings [@schruf2019; @cho2017; @xie2015; @bernard2015; @goodwin2018], we observed increases in both PER and OCR following 48 h of TGFβ stimulation, indicative of increased glycolysis and oxidative phosphorylation. Consistent with our direct measures of extracellular lactate and glucose, concurrent treatment with AZD3965 and VB124 was necessary to reduce PER. Consequently, this reduction in PER coincided with an increase in OCR as myofibroblasts transitioned their metabolism from glycolysis to oxidative phosphorylation.

@@

To further test this hypothesis, we performed liquid chromatography-mass spectrometry-based profiling of extracellular and intracellular metabolites from cells treated with AZD3965 and VB124. In accordance with the Seahorse analysis, inhibiting a single lactate transporter had modest effects on extracellular metabolite levels (**Fig. 4A-D, S4**). This analysis confirmed the results of extracellular lactate measurements by enzyme assay, which showed that dual inhibition was required to decrease lactate efflux (**Fig. 4B**). Besides lactate, dual inhibition of lactate transporters primarily altered transcellular fluxes of amino acids (**Fig. 4C-D**), including several metabolites that were differentially regulated by TGFβ treatment (**Fig. S4A**) and dual lactate transport inhibition, including leucine, alanine, ornithine, and ketoleucine.

Similar effects were observed on intracellular metabolites. PCA showed distinct clustering of treatment groups, with the drug effects aligning with PC1 and TGFβ treatment effects with PC2 (**Fig. 4E**). Similar to the extracellular flux results, the magnitude of drug-induced perturbations increased from MCT1 to MCT4 to combined inhibition (**Fig. 4E**), reflected by intracellular lactate levels (**Fig. 4F**). As expected from extracellular lactate measures following MCT inhibition, intracellular lactate accumulated moderately with MCT4 inhibition and substantially with MCT1/4 inhibition. TGFβ treatment resulted in diverse changes in the intracellular metabolomic profile of treated fibroblasts (**Fig. S4C-D**). The effect of AZD3965 alone on intracellular metabolite levels was modest (**Fig. S4G-H**). Conversely, MCT4 inhibition by VB124 alone (**Fig. SK-L**) or in combination with AZD3965 (**Fig. 4G-H**) caused substantial perturbations to intracellular metabolism. Specifically, we observed enrichment of the glycolysis and tricarboxylic acid (TCA) cycle metabolite sets with MCT4 inhibition. These findings collectively suggest that inhibiting lactate export leads to the accumulation of upstream glycolytic intermediates that are rerouted to mitochondrial oxidative metabolic pathways.

To test this hypothesis, we labeled lung fibroblasts with [U-^13^C~6~]-glucose (8 mM) in medium containing lactate (2 mM), glutamine (1 mM), and pyruvate (1 mM) during TGFβ stimulation and treatment with MCT inhibitors (**Fig. 4J, S5**). TGFβ increased ^13^C incorporation from glucose into pyruvate, lactate, citrate, succinate (SUC), and malate (MAL), indicating enhanced flux from glucose into the TCA cycle. While AZD3965 had minimal impact on these labeling patterns, MCT4 inhibition significantly elevated the fractions of these metabolites labeled by ^13^C, providing direct evidence for a proportional increase in glucose oxidation following MCT4 inhibition, consistent with the results of our bioenergetic and steady-state metabolomics experiments described above.

Increased metabolite labeling from [U-^13^C~6~]-glucose must be offset by decreased labeling from other substrates. Given recent data suggesting lactate as a major oxidative fuel source in the lung [@hui2017; @hui2020; @copeland2023], we hypothesized that MCT inhibition would decrease exogenous lactate oxidation. To test this, lung fibroblasts were cultured with [U-^13^C~3~]-lactate (2 mM) in medium containing naturally labeled glucose, glutamine, and pyruvate (**Fig. 4K, S5**). Extracellular [U-^13^C~3~]-lactate labeled approximately 50% of intracellular pyruvate and lactate at baseline, with significant downstream incorporation into TCA metabolites. This labeling decreased following TGFβ treatment, mirroring increased fractional labeling from glucose (**Fig. 4J**). MCT inhibition had no impact on fractional labeling of TCA intermediates by [U-^13^C~5~]-glutamine (**Fig. S4M, S5**). MCT4 inhibition alone or in combination with AZD3965 decreased ^13^C labeling of intracellular metabolites by lactate. These findings demonstrate the importance of MCT4 for lactate import at physiologic lactate concentrations, in contrast to the prevailing sentiment that MCT4 is primarily a lactate exporter. Moreover, these data further underscore the relatively greater importance of MCT4 activity in fibroblast metabolism compared to MCT1. Consistent with our measurements of extracellular lactate, the effects of MCT inhibition were more pronounced when both inhibitors were used simultaneously, again highlighting some functional redundancy of MCT1 and MCT4 in these cells.

@@

Given that increased intracellular lactate correlated with more potent inhibition of α-SMA expression, we treated cells with TGFβ in combination with extracellular lactate (10 mM). A previous study suggested that extracellular lactate modestly increases α-SMA expression independently of TGFβ [@kottmann2012]. However, we observed no impact of extracellular lactate on α-SMA expression in TGFβ-treated lung fibroblasts (**Fig. S7A**), indicating intracellular lactate accumulation alone does not mediate the antifibrotic effects of MCT inhibition.

@@

Our preclinical findings indicate that MCT4 inhibition, as a single therapeutic target, exhibits greater antifibrotic efficacy than MCT1 inhibition, both in suppressing myofibroblast differentiation *in vitro* and reducing bleomycin-induced fibrosis. VB253 is a novel inhibitor of MCT4 with ~10-fold increased selectivity for MCT4 *v*. MCT1 and ~10-fold increased potency for MCT4 inhibition. Like VB124, VB253 dose-dependently decreased TGFβ-stimulated α-SMA in human IPF lung fibroblasts *ex vivo* (**Fig. 8A**). To assess its efficacy compared to established therapies, we compared VB253 to nintedanib, a clinically approved antifibrotic medication for pulmonary fibrosis (**Fig. 8A-B**). Nintedanib inhibits TGFβ-mediated myofibroblast differentiation and decreases Col1a1 expression *in vitro* and *in vivo*  [@wollin2014]. While both compounds effectively attenuated α-SMA expression, nintedanib exhibited moderate cytotoxicity absent with VB253 (**Fig. 8B**). Nintedanib mitigates myofibroblast differentiation partly by inhibiting TGFβ receptor phosphorylation and Smad-dependent signaling pathways [@rangarajan2016]. Correspondingly, nintedanib dose-dependently attenuated TGFβ-mediated Smad3 phosphorylation, while VB253 and VB124 had no impact on this signaling pathway (**Fig. 5, S10**).

Subsequently, we evaluated the potential of VB253 to counter bleomycin-induced pulmonary fibrosis. In experiments conducted independently from those reported above, VB253 was administered to young mice (8-10 weeks) beginning on day 7 following bleomycin (**Fig. 8C**). The existing pulmonary fibrosis therapies, nintedanib and pirfenidone, served as comparators. Three weeks after bleomycin, whole-body plethysmography was performed to assess breathing patterns in unrestrained mice. Enhanced pause (Penh) is a dimensionless index describing airflow during tidal breathing found to increase significantly in bleomycin-treated mice [@vanoirbeek2010]. VB253 restored Penh to baseline levels, suggesting normalization of respiratory patterns in bleomycin-treated mice (**Fig. 8D**). Histologic assessment confirmed reduced pulmonary fibrosis (**Fig. 8E**) and α-SMA expression (**Fig. 8F**). Moreover, the effects of VB253 were comparable to the antifibrotic effects of nintedanib and pirfenidone with decreased cytotoxicity in *in vitro* assays.

@@

This study was designed to investigate the role of lactate transporters in myofibroblast differentiation and pulmonary fibrosis. The objectives of this study were (i) to determine the expression of lactate transporters in IPF lungs and model systems, (ii) to characterize the phenotypic effects of lactate transport inhibition in model systems, and (iii) to profile the metabolic consequences of lactate transport inhibitors.

Cell culture experiments were performed at least three times. The number of animals per experimental group was chosen based on prior publications and experiments were repeated at least once. Stable isotope tracing experiments were performed on three animals per group owing to resource availability. Animals were randomly assigned to treatment. The pathologist scoring histologic fibrosis severity was blinded to treatment assignment. The number of unique patient samples was determined by clinical availability.
